---
title: "Hypothesis tests"
version: 1.1
author: "Christina Bogner"
output:
  learnr::tutorial:
    progressive: true
    #css: [../../css/style_slides.css]
runtime: shiny_prerendered
description: "Introduction to hypothesis tests."
bibliography: [../../bib/statistik.bib]
biblio-style: apalike
link-citations: true
---

## Goal

Explain the idea behind simulation-based hypothesis tests


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(learnr)
library(emo)
library(tidyverse)
library(kableExtra)
library(infer)

a <- as.numeric(Sys.time())

study <- tibble('gender' = c(rep('male', 24),
                                        rep('female', 24)),
                      'decision' = c(rep('promoted', 21),
                                         rep('not_promoted', 3),
                                         rep('promoted', 14),
                                         rep('not_promoted', 10)))

real_diff <- table(study)[1,1]/sum(study$gender == 'male')*100 - table(study)[2,1]/sum(study$gender == 'female')*100
```

## Discrimination or pure chance? {.subsection}


A study by @Rosen1974 analysed whether 
**women are discriminated against in promotion decisions** [@Rosen1974].

- participants: 48 white male bank supervisors
- decide about promotion based on personal files (one by participant)
- personal files identical except gender
- random assignment of files to participants

<br>

This is an **experiment** 

`r emo::ji('right_arrow')` We can conclude on causality

### Data {.section}
```{r echo = FALSE}
knitr::kable(addmargins(table(study)), format = "html") %>%
    kable_styling(
      font_size = 15,
      bootstrap_options = c("striped", "hover", "condensed")
    ) 
```

### Research question
**Are women discriminated against in promotion decisions?**

<br>

Does the difference between men and women of `r round(real_diff, 2)`% mean that women are discriminated against, or is this difference due to chance?


### Which statement is correct?

- Statement 1: There is no real difference, the observed difference is due to chance.

H$_0$: **Null hypothesis**. The variables `gender` and `decision` are independent; the observed difference in promotions is random.

<br>

- Statement 2: Women are discriminated against.

H$_A$: **Alternative hypothesis**. The variables `gender` and `decision` are dependent. Women are discriminated against in decisions about promotion.

### Permute!
**Statement 1**: `gender` and `decision` are independent

`r emo::ji('right_arrow')` If we repeat the experiment, the difference will change because of random assignment of personal files only. 

`r emo::ji('right_arrow')` We toss the personal files and redistribute them again, but keep the decision fixed (i.e. we permute the variable `gender` and keep the variable `decision` as is!)

```{r echo = FALSE}
experiment <- sample(study$gender,
                     size = length(study$gender), replace = F)
```


```{r echo = FALSE}
actionButton(inputId = 'press',
             label = 'Go!')
textOutput("diff", inline = T)
htmlOutput('tab_experiment')
```


```{r, context="server"}
observeEvent(input$press,{
  experiment <- sample(study$gender, size = 
                         length(study$gender), replace = F)
  tab <- table(experiment, study$decision)[,1]
  n <- as.numeric(round(tab[1]/table(study$gender)[1]*100 - tab[2]/table(study$gender)[2]*100, 1))

    output$diff <- renderText(paste(' Difference:', n, '%'))

output$tab_experiment <- renderText({

  kable(addmargins(table(experiment, study$decision))) %>%
    kable_styling(
      font_size = 15,
      bootstrap_options = c("striped", "hover", "condensed")
    ) 
})
})                            

```


### Repeated permutation

```{r echo = FALSE}
sliderInput(inputId = "num", 
              label = "Number of permutations", 
              value = 1, min = 1, max = 10000)
```

```{r echo = FALSE}
plotOutput("daten", width = "60%")
```


```{r, context="server"}
output$daten = renderPlot({
res <- vector()
set.seed(123)
for (i in 1:input$num){
  experiment <- sample(study$gender, size = length(study$gender), replace = F)
  tab <- table(experiment, study$decision)[,1]
  res[i] <- tab[1]/table(study$gender)[1]*100 -
    tab[2]/table(study$gender)[2]*100
}

if (input$num <= 100) {
  par(bty = 'n')
  stripchart(res, offset = 0.5, vertical = F, xlim = c(-60, 60),
           method = 'stack', pch=19, ylim = c(0, 40), col = rgb(0,206,209, max = 255), cex = 2, xlab = 'Differences', 
           main = paste0('Proportion of differences > ', round(real_diff, 1), '%: ', round(sum(sapply(res, function(x) x >= real_diff))/length(res)*100, 2), '%'))
} else {
   hist(res, col = rgb(0,206,209, max = 255), xlab = 'Differences', ylab = 'Frequency',
        xlim = c(-60, 60), main = paste0('Proportion of differences > ', round(real_diff, 1), '%: ', 
round(sum(sapply(res, function(x) x >= real_diff))/length(res)*100, 2), '%'))
}
abline(v = real_diff, col = 'red', lwd = 2)
})

```

### Conclusion from permutations

- The difference of `r round(real_diff, 1)`% is rare.
- This contradicts the Statement 1 (**null hypothesis**) 
- We reject the **null hypothesis**.


## Hypothesis test

- Statistical method that tests whether an effect is due to chance.

- Null hypothesis: no effect statement

- Alternative hypothesis: the opposite of null, often the research question 

- If data contradicts the null hypothesis, it is rejected and the alternative accepted.


<br> 

Our example is a *permutation test*.

### $p$ value

How large is the proportion of differences of at least `r round(real_diff, 1)`%?

```{r echo = FALSE}
res <- vector()
set.seed(123)
for (i in 1:10000){
  experiment <- sample(study$gender, size = length(study$decision), replace = F)
  tab <- table(experiment, study$decision)[,1]
  res[i] <- tab[1]/table(study$gender)[1]*100 -
    tab[2]/table(study$gender)[2]*100
}

p_val <- round(sum(sapply(res, function(x) x >= real_diff))/length(res)*100, 2)
```

If we repeat the experiment 10000 times: `r p_val`%

**$p$ value** is the probability to obtain data as extreme or more extreme than what has been observed, if the null hypothesis is true.


<br>

For our example, it means:
Probability to obtain differences of at least `r round(real_diff, 1)`%, if the varialbes `gender` and `decision` are independent.

### Distribution of differences

- The distribution of differences changes with every new permutation of personal files
- It is a **random variable**
- with a symmetric distribution

```{r echo = F, out.width = "60%"}
hist(res, freq = F, breaks = seq(-60, 60, 10), col = rgb(0,206,209, max = 255), ylim = c(0, 0.04), main = 'Histogram of differences for 10000 permutations',
     ylab = 'Dichte', xlab = 'Differences')
lines(density(res, kernel = 'gaussian', bw = 'SJ', adjust = 75),
      col = 'orange', lwd = 3)
x = seq(-60, 60, 1)
lines(x, dnorm(x = x, mean = 0, sd = sd(res)), col = 'red', lwd = 3)
legend('topleft', legend = c('Probability density', 'Normal distribution'), col = c('orange', 'red'), lty = 1, bty = 'n')
```


### Why is the distribution symmetrical?

**Central limit theorem** [@Fahrmeir2016]: 
Proportions (or their differences) are normally distributed under the Null hypothesis. provided some mild assumptions are fulfilled.


```{r echo = F, out.width = "60%"}
hist(res, freq = F, breaks = seq(-60, 60, 10), col = rgb(0,206,209, max = 255), ylim = c(0, 0.04), main = 'Histogram of differences for 10000 permutations',
     ylab = 'Dichte', xlab = 'Differences')
lines(density(res, kernel = 'gaussian', bw = 'SJ', adjust = 75),
      col = 'orange', lwd = 3)
x = seq(-60, 60, 1)
lines(x, dnorm(x = x, mean = 0, sd = sd(res)), col = 'red', lwd = 3)
legend('topleft', legend = c('Probability density', 'Normal distribution'), col = c('orange', 'red'), lty = 1, bty = 'n')
```

### Conclusions from the permutation test

- $p$ value: probability to observe differences of at least `r round(real_diff, 1)`%, if the null hypothesis is true, i.e. the variables `gender` and `decision` are independent, equals `r p_val`%

- This is extremely rare.

- We reject the null hypothesis and conclude: Women were discriminated against in the 1970 (provided the white male banker are representative for the then-population of decision-makers).


## Literature

---
nocite: '@Diez2019'
...

